---
title: "Organization of gris tables"
author: "Michael Sumner"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(gris)
library(maptools)
```

The in-development R package `gris` provides a way to represent spatial objects as a set of relational tables. In short a `gris` object has tables "o" (objects), "b" (for branches), "bXv" (links between branches and vertices) and "v" the vertices. 

It can also have tables 'oXt' (links between objects and triangles) and 'tXv' (links between triangles and vertices) - this is not yet generalized to deal with line segments or multi-points rather than triangles so we ignore it for now. 



If we ingest the `wrld_simpl` layer we get a list with several tables. 

```{r}
library(gris)  ## devtools::install_github("mdsumner/gris")
library(maptools)
data(wrld_simpl)
gobject <- gris(wrld_simpl)
```
```{r,echo=FALSE}
gobject$bXv$.ob0 <- NULL
```

First the objects, these are individual countries with several attributes including the `NAME`. The attribute ".ob0" is a gris-specific one - the "object ID", there are also ".br0" attributes for branches, and ".vx0" for vertices. 

```{r}
gobject$o
```

The branches, these are individual simple, one-piece "ring polygons". Every object may have one or more branches (branches may be an "island" or a "hole" but this is not currently recorded). Note how branch 1 and 2 (`.br0`) both belong to object 1 (.ob0), but branch 3 is the only piece of object 2. 

```{r}
gobject$b

plot(gobject[1, ], col = "#333333")
title(gobject$o$NAME[1])
plot(gobject[2, ], col = "#909090")
title(gobject$o$NAME[2])
```

(Antigua and Barbuda sadly don't get a particularly good picture here, but this is not the point of the story.)

The links between branches and vertices.


```{r}
gobject$bXv
```


This table is required so that we can normalize the vertices by removing any duplicates based on X/Y pairs, a necessary preparation for the triangulation engine used belo (although not by the visualization). (Note that we could also normalize branches for objects, since multiple objects might use the same branch - but again off-topic). 

Finally, the vertices themselves. Here we only have X and Y, but these table structures can hold any number of attributes and of many types. 

```{r}
gobject$v
```
The normalization is only relevant for particular choices of vertices, so if we had X/Y/Z in use there might be a different version of "unique". I think this is a key point for flexibility, some of these tasks must be done on-demand and some ahead of time. 

Indices here are numeric, but there's actually no reason that they couldn't be character or other identifier.  Under the hood the `dplyr` package is in use for doing straightforward (and fast!) table manipulations including joins between tables and filtering on values. 

