---
title: "Polygons and coordinate systems"
author: "Michael Sumner"
date: "August 13, 2015"
output: html_document
---

Polygons have a veneer of sophistication as the underlying geometric and spatial tools required to work with them are quite advanced, but this masks the fact that polygons are quite simplistic and only suited to a narrow range of applications. Here we illustrate this problem and show some ways out. 

The simplest polygon is a piece of the plane bounded by three sides. 

```{r}
x <- c(-1, 1, 0)
y <- c(0, 0.2, 1)

xr <- range(x) + c(-1, 1) * diff(range(x))/4
yr <- range(y) + c(-1, 1) * diff(range(y))/4
plot(x, y, type = "n", xlim =xr, ylim = yr, asp = 1)
polypath(x, y)
```

In lay terms this shape is composed of 3 points joined by lines, and intuitively we can fill the inside of this shape as distinct from its outside and make explicit the three points. 

```{r}
plot(x, y, type = "n", xlim =xr, ylim = yr, asp = 1)
polypath(x, y, col = "grey")
points(x, y)
```

We can get confused about "points" here since to a mathematician this shape is composed of an infinity of points that fill the space inside the lines. To avoid this confusion we will use the term "vertex" rather than "point". Intuitively we need a vertex everywhere that our bounding lines change direction. Everywhere else along the line does not require an explicit vertex, since it's just a straight line.  

```{r}
plot(x, y, type = "n", xlim =xr, ylim = yr, asp = 1)
polypath(x, y, col = "grey")
points(x, y)
text(x, y, paste("Vertex", seq_along(x)), cex = 0.6, pos = c(1, 1, 3))
```

A triangle really is the most simple polygon, if we remove any vertex we only have a line, but another important property is that the triangle is **convex**.  If we add another vertex **on the existing boundary** and then move it there are really three kinds of outcomes. 

```{r}

x1 <- c(-1, 1, 0, 0)
y1 <- c(0, 0.2, 1, 0.6)
plot(x1, y1, main = "concave", xlim = xr, ylim =yr, asp = 1) 
polypath(x, y, col = rgb(0, 0, 0, 0.2), border = NA)
polypath(x1, y1, lty = 2)
text(x1, y1, paste("Vertex", seq_along(x1)), cex = 0.6, pos = c(1, 1, 3, 1))

x2 <- c(-1, 1, 0, -0.75)
y2 <- c(0, 0.2, 1, 0.6)
plot(x2, y2, main = "convex", xlim = xr, ylim =yr, asp = 1) 
polypath(x, y, col = rgb(0, 0, 0, 0.2), border = NA)
polypath(x2, y2, lty = 2)
text(x2, y2, paste("Vertex", seq_along(x2)), cex = 0.6, pos = c(1, 1, 3, 1))
x3 <- c(-1, 1, 0, -1/3)
y3 <- c(0, 0.2, 1, 2/3)
plot(x3, y3, main = "no change", xlim = xr, ylim =yr, asp = 1) 
polypath(x, y, col = rgb(0, 0, 0, 0.2), border = NA)
polypath(x3, y3, lty = 2)
text(x3, y3, paste("Vertex", seq_along(x3)), cex = 0.6, pos = c(1, 1, 3, 1))
```

These outcomes aren't all guaranteed though, as there is two kinds of ambiguity here. 

Consider again the case where our new vertex is internal to the original triangle. This might be a polygon with a concave left side, right side, or lower side based on which ever pair of the outer vertices is put with the inner vertex. 


```{r}
op <- par(mfrow = c(2, 2))
x1 <- c(-1, 1, 0, 0)
y1 <- c(0, 0.2, 1, 0.6)
plot(x1, y1, main = "one concavity", xlim = xr, ylim =yr, asp = 1) 
polypath(x, y, col = rgb(0, 0, 0, 0.1), border = NA)
polypath(x1, y1, lty = 2)
text(x1, y1, paste("Vertex", seq_along(x1)), cex = 0.6, pos = c(1, 1, 3, 1))

plot(x1, y1, main = "two concavity", xlim = xr, ylim =yr, asp = 1) 
polypath(x, y, col = rgb(0, 0, 0, 0.1), border = NA)
polypath(cbind(x1, y1)[c(1, 2, 4, 3), ], lty = 2)
text(x1, y1, paste("Vertex", seq_along(x1)), cex = 0.6, pos = c(1, 1, 3, 1))

plot(x1, y1, main = "three concavity", xlim = xr, ylim =yr, asp = 1) 
polypath(x, y, col = rgb(0, 0, 0, 0.1), border = NA)
polypath(cbind(x1, y1)[c(1, 4, 2, 3), ], lty = 2)
text(x1, y1, paste("Vertex", seq_along(x1)), cex = 0.6, pos = c(1, 1, 3, 1))
par(op)
```

Each of these polygons is very different, and each is actually composed of a different set of underlying triangles. 



Winding and even-odd rule. 

```{r}
op <- par(mfrow = c(2, 2))
v1 <- cbind(x = c(0, 1, 0.5), y = c(0, 0, 1))
v0 <- cbind(x = c(0.4, 0.6, 0.5) - 0.3, y = c(0.4, 0.4, 0.6))

plot(v1, main = "winding, counterclockwise")
points(v0)
polypath(rbind(v1, NA,  v0), col = "grey", rule = "winding")

plot(v1, main = "winding, clockwise")
points(v0)
polypath(rbind(v1, NA,   v0[nrow(v0):1, ]), col = "grey", rule = "winding")


plot(v1, main = "evenodd, counterclockwise")
points(v0)
polypath(rbind(v1, NA,  v0), col = "grey", rule = "evenodd")

plot(v1, main = "evenodd, clockwise")
points(v0)
polypath(rbind(v1, NA,   v0[nrow(v0):1, ]), col = "grey", rule = "evenodd")
par(op)
```